% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother


\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal Name}

\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={CLEM-ICD - Clinical Language Encoding with ModernBERT},
  pdfauthor={Tyler Cross},
  pdfkeywords={ModernBERT, natural language
processing, MIMIC-IV, ICD-10},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{CLEM-ICD - Clinical Language Encoding with
ModernBERT \\\large{Using ModernBERT to Improve Automated ICD-10
Classification} }
\author[1]{Tyler Cross%
\corref{cor1}%
\fnref{fn1}}
 \ead{tyler.cross@berkeley.com} 

\affiliation[1]{organization={University of California, Berkeley, School
of Information},addressline={102 South
Hall},city={Berkeley},postcode={94720},postcodesep={}}

\cortext[cor1]{Corresponding author}
\fntext[fn1]{Final project submission for the UC Berkeley Master of
Information and Data Science program, DATASCI 266: Natural Language
Processing with Deep Learning}
        
\begin{abstract}
Medical coding is a critical yet complex process that requires
translating clinical documentation into standardized ICD-10 codes.
Currently, healthcare providers rely on dedicated medical coders who
manually review extensive clinical notes to assign appropriate codes
from 150,000+ possible diagnostic and procedural codes. This project
builds upon the work of \citet{edin2024explainable}{]} with two key
innovations: (1) replacing RoBERTa with ModernBERT
\citep{warner2024modernbert} as the encoder to potentially improve
performance through its enhanced context window, and (2) adding a
LLM-powered reasoning layer that provides natural language explanations
for code predictions, bridging the gap between model predictions and
human validation. Using the MIMIC-IV dataset, we implement and evaluate
this approach using standard metrics (precision, recall, F1) and
explanation quality.
\end{abstract}





\begin{keyword}
    ModernBERT \sep natural language processing \sep MIMIC-IV \sep 
    ICD-10
\end{keyword}
\end{frontmatter}
    

Medical coding is a critical yet complex process in healthcare
administration. The translation of clinical documentation into
standardized ICD-10 codes is essential for healthcare reimbursement,
clinical research, and public health statistics. Currently, healthcare
providers rely on dedicated medical coders who manually review extensive
clinical notes to assign appropriate codes from over 150,000 possible
diagnostic and procedural codes, creating significant administrative
overhead.

For every \$1 of revenue collected by a hospital, approximately \$0.25
is spent on administrative tasks necessary to collect it. Non-doctor
workers outnumber doctors in healthcare roughly 16 to 1, with medical
coding representing a significant portion of this burden.

This project aims to improve automated medical coding by leveraging
ModernBERT's enhanced context window capabilities, potentially reducing
the administrative burden while maintaining or improving coding
accuracy.

\section{Background}\label{background}

\citet{edin2024explainable} recently demonstrated success in medical
coding using a RoBERTa-based approach. Their work showed promising
results but left room for improvement in both performance and contextual
understanding. Other notable contributions in this field include
{[}additional related work to be filled in{]}.

ModernBERT \citep{warner2024modernbert} represents an advancement over
previous BERT-like models, offering an enhanced context window of 8192
tokens compared to the 512 tokens in traditional models. This increased
context capacity is particularly valuable for medical documents, which
tend to be lengthy and contain important information distributed
throughout the text.

\subsection{Using CSL}\label{using-csl}

\section{Methods}\label{methods}

This project utilizes the MIMIC-IV dataset, a large, freely available
database comprising de-identified health data associated with hospital
stays. The dataset includes detailed clinical notes, diagnostic codes,
procedural information, and other health-related data from real hospital
encounters.

\subsection{System Architecture}\label{system-architecture}

Our approach replaces the RoBERTa model used in previous work
\citep{edin2024explainable} with ModernBERT
\citep{warner2024modernbert}, leveraging its enhanced context window to
better process lengthy clinical notes. We hypothesize that this will
lead to improved code prediction by enabling the model to capture more
comprehensive context from medical documents.

\subsection{Implementation Details}\label{implementation-details}

{[}To be completed with specific implementation details{]}

\subsection{Evaluation Approach}\label{evaluation-approach}

We evaluate our system using standard performance metrics: precision,
recall, F1-score, and accuracy compared to gold-standard human coding,
following the evaluation protocol established in previous work.

\section{Results and Discussion}\label{results-and-discussion}

{[}This section will be completed after obtaining experimental
results{]}


\renewcommand\refname{References}
  \bibliography{bibliography.bib}



\end{document}
